---
title: "CHL8010: Armed Conflict - Exploratory Data Analysis"
author: "Mark Asuncion"
date: "2024-09-30"
format: 
  pdf:
    number-sections: true  # Number the sections
    margin-left: 1in    # Adjust margins as needed
    margin-right: 1in
    margin-top: 1in
    margin-bottom: 1in
    keep-tex: true      # Keep the intermediate .tex file for inspection
fontsize: 11pt          # Set the font size
geometry: "a4paper"     # Paper size (adjust if needed)
---

```{r, include=FALSE}
options(warn = -1)
library(here)
library(tidyverse)
library(cowplot)
library(ggplot2)
finaldata <- read.csv(here("original", "primary_data.csv"), row.names = NULL)
```

This file serves to document the exploratory data analysis for the merged data that
was compiled last week.

# Viewing the Data
Using `head()` and `tail()` gives us the data for the first/last countries
```{r, echo=TRUE}
head(finaldata, 20)
tail(finaldata, 20)
```
From first glance, one prominent feature of the data is that several variables
have missing values - which gives us an idea of what to explore first when the
time comes to deal with this. Moreover we can confirm that the values are conforming
to what they should be by defintion. We also can confirm that the columns we 
created for the disaster/conflict data have had their NA values imputed with 0s
respectively.

# Gathering Summary Statistics
For the numerical variables, we will be looking at their summary statistics and 
checking to see if the removal of certain data points (or countries) leads to a 
drastic change. First we can check the overall summary statistics:
```{r, echo=TRUE}
# Capturing all numerical variables
numeric_dat <- finaldata[sapply(finaldata, is.numeric)]
print(summary(numeric_dat))
```
Next we can look at certain countries like Canada, for example:
```{r, echo=TRUE}
# Capturing all numerical variables for Canada
canada_dat <- subset(finaldata, country_name == "Canada")
numeric_dat <- canada_dat[sapply(canada_dat, is.numeric)]
print(summary(numeric_dat))
```
One thing we can explore is the distribution of the total number of conflicts 
per year, from there we can look further at the countries on the extremes.
```{r, echo=TRUE}
years = unique(finaldata$year)
for (y in years) {
  dat <- subset(finaldata, year == y)
  print(sprintf("Summary of total conflicts for year %d", y))
  print(summary(dat$conflict))
}
```
Comparing to the overall summary, we can see that 2015 is where the max number
of conflicts occurred (the minimum occurs across several years). Further we can
match this to the country, specifically this occurred in Syria.

```{r, echo=TRUE}
subset(finaldata, conflict >= 78000) # The max number of conflicts happened in Syria during 2015
```

# Data Visualization
In this section we aim to explore any relationships between variables by using 
different visuals. Given the most number of deaths occurred in Syria, let's 
visualize any trends in this country over the years.
```{r, echo=TRUE}
countries <- unique(finaldata$country_name)
data = subset(finaldata, country_name == "Syria")

# Number of Conflicts vs Year in Syria
ggplot(data, aes(x = year, y = conflict)) +
      geom_line(color = "blue") + 
      geom_point(color = "red") + 
      labs(title = "Conflict in Syria over the years", x = "Year", 
           y = "Total Number of Conflicts") 

# Reshape the data to long format
syria = subset(finaldata, country_name == "Syria")
data_long <- syria %>%
  pivot_longer(cols = c(MatMor, InfMor, Und5Mor, NeoMor), 
               names_to = "MortalityType", 
               values_to = "Rate")

# Create the plot
ggplot(data_long, aes(x = year, y = Rate, color = MortalityType)) +
      geom_line() +
      geom_point() + 
      labs(title = "Mortality Rates in Syria over the years", 
           x = "Year", 
           y = "Mortality Rate") +
      scale_color_manual(values = c("MatMor" = "blue", 
                                      "InfMor" = "red", 
                                      "Und5Mor" = "green", 
                                      "NeoMor" = "orange"))
```
These two plots align with each other in the fact that their maximums occur around
2015, which is expected. When there was a higher number of conflicts, the mortality
rates (perhaps excluding neonatal) also obtained their max values in Syria.